On the surprising effectiveness of a simple matrix exponential derivative approximation, with application to global SARS-CoV-2

### 2023-09-14 manuscript/pnas_revision.tex
fixed numbering issues

 \newtheorem{theorem}{\bf Theorem}
 \newtheorem{lemma}{\bf Lemma}
 \newtheorem{fact}{\bf Fact}
 \newtheorem{corollary}{\bf Corollary}
 \newtheorem{axiom}{\bf Axiom}
 \newtheorem{condition}{\bf Condition}
 \newtheorem{property}{\bf Property}
 \newtheorem{proposition}{\bf Proposition}

@@

These observations lead to the following theorem, the proof of which appears
in the supplement as \cref{sec:Thm:det:main}.
\begin{Theorem}\label{thm:ass:error}
  Suppose that $\QQ, \JJ \in \mathcal{M}(d)$ for some $d \geq 1$.  We
  assume that we can find an element $\QQ^+ \in \mathcal{M}(d)$ such that
  $\QQ$ is a generalized inverse of $\QQ^+$, namely,
  \begin{align}\label{eq:GI:prop}
    \QQ^+ \QQ \QQ^+ = \QQ^+
  \end{align}
  and such that
  \begin{align}\label{eq:MP:proj:prop}
    e^{\tau \QQ}(\II - \QQ^+ \QQ) \!=\!  \II - \QQ^+ \QQ, \;
    (\II - \QQ \QQ^+)e^{\tau \QQ} \!=\!  \II - \QQ \QQ^+,
  \end{align}
  for any $\tau \in \bbR$. Furthermore, we suppose that $\QQ^+$ and
  $\QQ$ commute
  \begin{align}\label{eq:GI:com}
    \QQ \QQ^+ = \QQ^+ \QQ.
  \end{align}
  Finally, taking $\|\cdot\|$ be any matrix norm, we assume that
  \begin{align}\label{eq:qetq:exp:decay}
    \| \QQ e^{\tau \QQ}\| \leq C_0 e^{-\kappa \tau},
    \quad
    \text{ for all } \tau \geq 0,
  \end{align}
  where the constants $C_0 > 0, \kappa > 0$
  are independent of $\tau$.  Then, under these circumstances, 
  \begin{align}\label{eq:ass:error}
    \|\QQ^+ \JJ (\II - \QQ \QQ^{+})-& (\II -\QQ^+ \QQ) \JJ \QQ^{+}
        +t(\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+}
                \notag\\
      &
        +  \nabla_\JJ e^{t \QQ} - t  e^{t\QQ} \JJ\| \leq C(1+ t)e^{-\kappa t}
\end{align}
for any $t \geq 0$.  Here, $C > 0$ is a $t$-independent
constant which is given explicitly as
\begin{align}
  C_0 \big(\| (\II -\QQ^+ \QQ) \JJ (\QQ^{+})^2\|
  \!&+\! \|(\QQ^+)^2\JJ (\II - \QQ \QQ^{+})\|
    \!+\! \|\QQ^+ \hspace{0.5mm}\JJ\|\big)
    \notag\\
    &+ C_0^2 \| \QQ^+  \JJ  \QQ^{+} \|.
      \label{eq:exp:error:ass:er}
  \end{align}
\end{Theorem}

\begin{Remark}\label{rmk:mat:class:dcy}

@@

\end{Remark}

@@

\begin{Theorem}\label{thm:bride:of:frankinstein}

@@

\end{Theorem}

@@

We formulate our second major result as follows. 
\begin{Theorem}\label{t:lambda_n-1(Q)->-infty_non-symm_a.s.}
  Let $F_X \in L_{\psi_1}$ be a distribution such that $X \geq 0$
  a.s., $\bbE X = \mu > 0$ and $\Var X = \sigma^2 > 0$. We consider a
  sequence of random matrices
  $\QQ \equiv \QQ(d) = \{q_{ij}\}_{i,j=1,\hdots,d} \in \mathcal{R}(d)$, $d \in \bbN$,
  where either
  \begin{align}\label{e:RM_theorem_general_condition}
    \begin{aligned}
  &\{q_{ij}\}_{i , j =1, \hdots, d, \hspace{0.75mm}i \neq j}
    \stackrel{\textnormal{iid}}\sim F_X
    \text{ and }\\
  &- q_{ii} = \sum_{j \in \{1,\hdots,d\} \backslash \{i\}}q_{ij},
  i=1,\hdots,d
  \end{aligned}
  \end{align}
  or we impose that $\QQ \in \mathcal{R}(d) \cap \mathcal{S}(d, \RR)$ as 
    \begin{align}
    \begin{aligned}\label{e:RM_theorem_symmetry_condition}
  &\{q_{ij}\}_{i , j =1, \hdots, d, \hspace{0.75mm}i > j}
  \stackrel{\textnormal{iid}}\sim F_X,
  q_{ij} := q_{ji} \text{ for } i < j
    \text{ and }\\
  &- q_{ii} = \sum_{j \in \{1,\hdots,d\} \backslash \{i\}}q_{ij},
  i=1,\hdots,d.
  \end{aligned}
\end{align}
    Then, in either of these cases, for any $d \in \bbN \backslash\{1\}$, we have
    \begin{align}
 \mu+ O_{\as}\Big(\sqrt{ \frac{\log d}{d}}\Big) \leq& \frac{\sigma_2(\QQ)}{d} 
 \notag \\
 \leq& \frac{\sigma_d(\QQ)}{d}
 \leq \mu+ O_{\as}\Big(\sqrt{ \frac{\log d}{d}}\Big)
 \label{e:lambda_n-1(Q)->-infty_non-symm}
      \end{align}
      almost surely. In \eqref{e:lambda_n-1(Q)->-infty_non-symm},
      $O_{\textnormal{a.s.}}$ is as in
      \eqref{e:X=O_a.s.(1)_Y=o_a.s.(1)}.
\end{Theorem}

@@

\begin{Remark}\label{rmk:non:sym:num:ev}
  Our rigorous formulation of \cref{cor:rig:det:rand:wrapup} is
  limited to symmetric random rate matrices whose above diagonal
  elements are independent and identically distributed draws from a
  sub-exponential distribution. However, strong numerical evidence
  suggest that the scope of the approximations
  \eqref{eq:bride:of:frankinstein:1:rm:app},
  \eqref{eq:bride:of:frankinstein:0:rm:app} reach far beyond
  limitations of \cref{cor:rig:det:rand:wrapup} in several different
  ways. \cref{fig:app} explores the
  consequences of relaxing various assumptions of
  \cref{cor:rig:det:rand:wrapup}. In general, there appears to be no
  significant departure from the idea that the proxy form
  $t(\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+}$ corresponds increasingly well to
  the true approximation error (\eqref{eq:true:diff}) as the dimension
  increases.
\end{Remark}

\begin{Remark}\label{rmk:better:approx}
  Calculation of
  $\widetilde{\nabla}_{\JJ} e^{t\QQ} := t e^{t\QQ} \JJ$ requires
  $\order{d^3}$ operations by, e.g., computing the spectral
  decomposition $\QQ$ as in \eqref{eq:spec:decomp}.  One may then recycle this
  decomposition to determine the
  additional term $t(\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+}$ for little extra cost.  In view of
  \cref{cor:rig:det:rand:wrapup} and \cref{fig:app},
  \begin{align}\label{eq:better:help:on:sale}
    \widehat{\nabla}_{\JJ} e^{t\QQ} := t e^{t\QQ} \JJ
    -t(\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+}
  \end{align}
  provides an accurate approximation of $\nabla_{\JJ} e^{t\QQ}$ for
  asymptotically for large $d$. Thus, we anticipate further computational improvements when fitting
  large, gold-standard models using this
  refined approximation.  That said, we leave the efficient and scalable application of $\widehat{\nabla}_{\JJ} e^{t\QQ}$ to future work.
\end{Remark}

>>>>

% \newtheorem{remark}{\bf Remark}
 \newtheorem{theorem}{\bf Theorem}
 \newtheorem{lemma}{\bf Lemma}
 \newtheorem{fact}{\bf Fact}
 \newtheorem{corollary}{\bf Corollary}
 \newtheorem{axiom}{\bf Axiom}
 \newtheorem{condition}{\bf Condition}
 \newtheorem{property}{\bf Property}
 \newtheorem{proposition}{\bf Proposition}

@@

These observations lead to the following theorem, the proof of which appears
in the supplement as \cref{sec:Thm:det:main}.
\begin{theorem}\label{thm:ass:error}
  Suppose that $\QQ, \JJ \in \mathcal{M}(d)$ for some $d \geq 1$.  We
  assume that we can find an element $\QQ^+ \in \mathcal{M}(d)$ such that
  $\QQ$ is a generalized inverse of $\QQ^+$, namely,
  \begin{align}\label{eq:GI:prop}
    \QQ^+ \QQ \QQ^+ = \QQ^+
  \end{align}
  and such that
  \begin{align}\label{eq:MP:proj:prop}
    e^{\tau \QQ}(\II - \QQ^+ \QQ) \!=\!  \II - \QQ^+ \QQ, \;
    (\II - \QQ \QQ^+)e^{\tau \QQ} \!=\!  \II - \QQ \QQ^+,
  \end{align}
  for any $\tau \in \bbR$. Furthermore, we suppose that $\QQ^+$ and
  $\QQ$ commute
  \begin{align}\label{eq:GI:com}
    \QQ \QQ^+ = \QQ^+ \QQ.
  \end{align}
  Finally, taking $\|\cdot\|$ be any matrix norm, we assume that
  \begin{align}\label{eq:qetq:exp:decay}
    \| \QQ e^{\tau \QQ}\| \leq C_0 e^{-\kappa \tau},
    \quad
    \text{ for all } \tau \geq 0,
  \end{align}
  where the constants $C_0 > 0, \kappa > 0$
  are independent of $\tau$.  Then, under these circumstances, 
  \begin{align}\label{eq:ass:error}
    \|\QQ^+ \JJ (\II - \QQ \QQ^{+})-& (\II -\QQ^+ \QQ) \JJ \QQ^{+}
        +t(\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+}
                \notag\\
      &
        +  \nabla_\JJ e^{t \QQ} - t  e^{t\QQ} \JJ\| \leq C(1+ t)e^{-\kappa t}
\end{align}
for any $t \geq 0$.  Here, $C > 0$ is a $t$-independent
constant which is given explicitly as
\begin{align}
  C_0 \big(\| (\II -\QQ^+ \QQ) \JJ (\QQ^{+})^2\|
  \!&+\! \|(\QQ^+)^2\JJ (\II - \QQ \QQ^{+})\|
    \!+\! \|\QQ^+ \hspace{0.5mm}\JJ\|\big)
    \notag\\
    &+ C_0^2 \| \QQ^+  \JJ  \QQ^{+} \|.
      \label{eq:exp:error:ass:er}
  \end{align}
\end{theorem}

\begin{remark}\label{rmk:mat:class:dcy}

@@

\end{remark}

@@

\begin{theorem}\label{thm:bride:of:frankinstein}

@@

\end{theorem}

@@

We formulate our second major result as follows. 
\begin{theorem}\label{t:lambda_n-1(Q)->-infty_non-symm_a.s.}
  Let $F_X \in L_{\psi_1}$ be a distribution such that $X \geq 0$
  a.s., $\bbE X = \mu > 0$ and $\Var X = \sigma^2 > 0$. We consider a
  sequence of random matrices
  $\QQ \equiv \QQ(d) = \{q_{ij}\}_{i,j=1,\hdots,d} \in \mathcal{R}(d)$, $d \in \bbN$,
  where either
  \begin{align}\label{e:RM_theorem_general_condition}
    \begin{aligned}
  &\{q_{ij}\}_{i , j =1, \hdots, d, \hspace{0.75mm}i \neq j}
    \stackrel{\textnormal{iid}}\sim F_X
    \text{ and }\\
  &- q_{ii} = \sum_{j \in \{1,\hdots,d\} \backslash \{i\}}q_{ij},
  i=1,\hdots,d
  \end{aligned}
  \end{align}
  or we impose that $\QQ \in \mathcal{R}(d) \cap \mathcal{S}(d, \RR)$ as 
    \begin{align}
    \begin{aligned}\label{e:RM_theorem_symmetry_condition}
  &\{q_{ij}\}_{i , j =1, \hdots, d, \hspace{0.75mm}i > j}
  \stackrel{\textnormal{iid}}\sim F_X,
  q_{ij} := q_{ji} \text{ for } i < j
    \text{ and }\\
  &- q_{ii} = \sum_{j \in \{1,\hdots,d\} \backslash \{i\}}q_{ij},
  i=1,\hdots,d.
  \end{aligned}
\end{align}
    Then, in either of these cases, for any $d \in \bbN \backslash\{1\}$, we have
    \begin{align}
 \mu+ O_{\as}\Big(\sqrt{ \frac{\log d}{d}}\Big) \leq& \frac{\sigma_2(\QQ)}{d} 
 \notag \\
 \leq& \frac{\sigma_d(\QQ)}{d}
 \leq \mu+ O_{\as}\Big(\sqrt{ \frac{\log d}{d}}\Big)
 \label{e:lambda_n-1(Q)->-infty_non-symm}
      \end{align}
      almost surely. In \eqref{e:lambda_n-1(Q)->-infty_non-symm},
      $O_{\textnormal{a.s.}}$ is as in
      \eqref{e:X=O_a.s.(1)_Y=o_a.s.(1)}.
\end{theorem}

@@

\begin{remark}\label{rmk:non:sym:num:ev}
  Our rigorous formulation of \cref{cor:rig:det:rand:wrapup} is
  limited to symmetric random rate matrices whose above diagonal
  elements are independent and identically distributed draws from a
  sub-exponential distribution. However, strong numerical evidence
  suggest that the scope of the approximations
  \eqref{eq:bride:of:frankinstein:1:rm:app},
  \eqref{eq:bride:of:frankinstein:0:rm:app} reach far beyond
  limitations of \cref{cor:rig:det:rand:wrapup} in several different
  ways. \cref{fig:app} explores the
  consequences of relaxing various assumptions of
  \cref{cor:rig:det:rand:wrapup}. In general, there appears to be no
  significant departure from the idea that the proxy form
  $t(\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+}$ corresponds increasingly well to
  the true approximation error (\eqref{eq:true:diff}) as the dimension
  increases.
\end{remark}

\begin{remark}\label{rmk:better:approx}
  Calculation of
  $\widetilde{\nabla}_{\JJ} e^{t\QQ} := t e^{t\QQ} \JJ$ requires
  $\order{d^3}$ operations by, e.g., computing the spectral
  decomposition $\QQ$ as in \eqref{eq:spec:decomp}.  One may then recycle this
  decomposition to determine the
  additional term $t(\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+}$ for little extra cost.  In view of
  \cref{cor:rig:det:rand:wrapup} and \cref{fig:app},
  \begin{align}\label{eq:better:help:on:sale}
    \widehat{\nabla}_{\JJ} e^{t\QQ} := t e^{t\QQ} \JJ
    -t(\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+}
  \end{align}
  provides an accurate approximation of $\nabla_{\JJ} e^{t\QQ}$ for
  asymptotically for large $d$. Thus, we anticipate further computational improvements when fitting
  large, gold-standard models using this
  refined approximation.  That said, we leave the efficient and scalable application of $\widehat{\nabla}_{\JJ} e^{t\QQ}$ to future work.
\end{remark}

### 2023-09-14 manuscript/revision_letter.tex
fixed numbering issues

This is a review for the manuscript "On the surprising effectiveness of a simple matrix exponential derivative approximation, with application to global SARS-CoV-2". 
For the benefit of the authors on how they interpret my review, I am a theoretical/computational statistician with background on theoretical population genetics, interested in Markov chains and its implementations in Monte Carlo methods.

@@

@@

@@

>>>>

This is a review for the manuscript ``On the surprising effectiveness of a simple matrix exponential derivative approximation, with application to global SARS-CoV-2''. 
For the benefit of the authors on how they interpret my review, I am a theoretical/computational statistician with background on theoretical population genetics, interested in Markov chains and its implementations in Monte Carlo methods.

\begin{reply}
	We thank you for lending your effort and expertise in reviewing our manuscript and for your helpful recommendations.  We have thoroughly implemented your suggestions and believe that the manuscript is greatly improved.  
\end{reply}

@@

\begin{reply}
	We thank you for catching this numbering error, which we have addressed.  Now there are 3 theorems (1,2,3) and 5 remarks (1,2,3,4,5).
\end{reply}

@@

\begin{reply}
	We thank you for your positive reception of our mathematical work.
\end{reply}

@@

\begin{reply}
	We thank you for sharing these concerns.  Before responding, let us point out that we have now added a simulation section (similar to that which you specify below) to the manuscript.   Now, we would like to argue that the decisions we made in structuring our manuscript are well-motivated based on our work in two other papers.
	\begin{enumerate}
		\item First,
	\end{enumerate}
\end{reply}
